Why there are more species in several small patches vs few large patches:

### 2023-08-22 RunSims.Rmd
Update manuscript

# Neutral communities 
#
mdl <- sloss_sim %>% filter(deforestation_type == "regular",birds_behavior=="NoSelection")
names(mdl)
#
# Plot for neutral communities
# Calculate the power exponent 
# pexp = (1 - 2 * birds_dispersal_distance ) / (1 - birds_dispersal_distance ), median_dispersal_distance = 2^(1/pexp-1) 
#
mdl1 <- mdl %>% mutate(habitat_patch_size = habitat_patch_size+1) %>%
                         group_by(birds_dispersal_distance, habitat_patch_size,birth_rate_birds) %>% summarise(across(c(birds_at_deforestation:count_birds,max_patch),list(mean=mean, sd=sd, ql=~quantile(.x,probs=c((1-0.95)/2)), qh=~quantile(.x,probs=c(1-(1-0.95)/2)) )))

@@

```{r Selection_SLOSS_lambda14_simulations_regular,echo=FALSE,message=FALSE,warning=FALSE,eval=TRUE}

@@

## Not fragmented communities Lambda = 1.7 - 4

>>>>

# Neutral communities 
#
mdl <- sloss_sim %>% filter(deforestation_type == "regular",birds_behavior=="NoSelection", habitat_patch_size != 199)
names(mdl)
#
# Plot for neutral communities
# Calculate the power exponent 
# pexp = (1 - 2 * birds_dispersal_distance ) / (1 - birds_dispersal_distance ), median_dispersal_distance = 2^(1/pexp-1) 
#
mdl1 <- mdl %>% mutate(habitat_patch_size = habitat_patch_size+1) %>%
                         group_by(birds_dispersal_distance, habitat_patch_size,birth_rate_birds) %>% summarise(across(c(birds_at_deforestation:count_birds,max_patch),list(mean=mean, sd=sd, ql=~quantile(.x,probs=c((1-0.95)/2)), qh=~quantile(.x,probs=c(1-(1-0.95)/2)) )))

@@

```{r Selection_SLOSS_lambda14_simulations_regular,echo=FALSE,message=FALSE,warning=FALSE,eval=FALSE}

@@

# Plot for birth selection communities regular patterns

```{r ExploratoryPlotsSelectionRegular, echo=FALSE,message=FALSE,warning=FALSE,fig.align='center',eval=FALSE}

mdl <- sloss_sim %>% filter(deforestation_type == "regular", birds_behavior=="BirthSelection")
names(mdl)
#
# summarize data 
#
mdl1 <- mdl %>% mutate(habitat_patch_size = habitat_patch_size+1) %>%
                         group_by(birds_dispersal_distance, habitat_patch_size,birth_rate_birds) %>% summarise(across(birds_at_deforestation:count_birds,list(mean=mean, sd=sd, ql=~quantile(.x,probs=c((1-0.95)/2)), qh=~quantile(.x,probs=c(1-(1-0.95)/2)) ))) #  %>% mutate(deltaPop = (count_birds - birds_at_deforestation )/ birds_at_deforestation  )
names(mdl1)

ggsave("Figures/Selection_Species_Dispersal_Lambda_regular.png",width=8,height=6,units="in",dpi=600)

# Shannon-Diversity  
#
mdl1 %>%  
  ggplot(aes(x = habitat_patch_size, y = calc_shannon_diversity_mean, color = factor(birds_dispersal_distance))) + 
  geom_line(aes(group = interaction(birds_dispersal_distance, birth_rate_birds)),  size = 1.2) +
  geom_ribbon(aes(ymin = calc_shannon_diversity_ql, ymax = calc_shannon_diversity_qh, group = interaction(birds_dispersal_distance, birth_rate_birds), fill = factor(birds_dispersal_distance)), alpha = 0.3,colour=NA, show.legend = FALSE) +
  scale_color_viridis_d() +
  facet_wrap(~ birth_rate_birds, ncol = 4) + 
  labs(color = "Dispersal\n distance") +
  theme_bw() + xlab("Habitat Patch Size") + ylab("Shannon Diversity") +
  theme(
    legend.position = c(0.9,0.8)
    )

ggsave("Figures/Selection_Shannon_Dispersal_Lambda_regular.png",width=8,height=6,units="in",dpi=600)

## Plots across neutral hierarchical selection modelos regular fragmentation

```{r PlotsAcrossBehaviorRegular, echo=FALSE,message=FALSE,warning=FALSE,fig.align='center',eval=FALSE}

mdl <- sloss_sim %>% filter(deforestation_type == "regular")
names(mdl)
#
# summarize data 
#
mdl1 <- mdl %>% mutate(habitat_patch_size = habitat_patch_size+1) %>%
                         group_by(birds_dispersal_distance, habitat_patch_size,birth_rate_birds,birds_behavior) %>% summarise(across(birds_at_deforestation:count_birds,list(mean=mean, sd=sd, ql=~quantile(.x,probs=c((1-0.95)/2)), qh=~quantile(.x,probs=c(1-(1-0.95)/2)) ))) #  %>% mutate(deltaPop = (count_birds - birds_at_deforestation )/ birds_at_deforestation  )
names(mdl1)

mdl1 %>%  filter(birth_rate_birds == 2) %>% 
  ggplot(aes(x = habitat_patch_size, y = calc_number_of_species_mean, color = factor(birds_dispersal_distance))) + 
  geom_line(aes(group = interaction(birds_dispersal_distance, birds_behavior)),  size = 1.2) +
  geom_ribbon(aes(ymin = calc_number_of_species_ql, ymax = calc_number_of_species_qh, group = interaction(birds_dispersal_distance, birds_behavior), fill = factor(birds_dispersal_distance)), alpha = 0.3,colour=NA, show.legend = FALSE) +
  scale_color_viridis_d() +
  facet_wrap(~ birds_behavior, ncol = 4) + 
  labs(color = "Dispersal\n distance") +
  theme_bw() + xlab("Habitat Patch Size") + ylab("Number of species") +
  theme(
    legend.position = c(0.9,0.8)
    )

ggsave("Figures/AcrossBehav_Species_Dispersal_Lambda2_regular.png",width=8,height=6,units="in",dpi=600)

# Shannon-Diversity  
#
mdl1 %>%  filter(birth_rate_birds == 2) %>% 
  ggplot(aes(x = habitat_patch_size, y = calc_shannon_diversity_mean, color = factor(birds_dispersal_distance))) + 
  geom_line(aes(group = interaction(birds_dispersal_distance, birds_behavior)),  size = 1.2) +
  geom_ribbon(aes(ymin = calc_shannon_diversity_ql, ymax = calc_shannon_diversity_qh, group = interaction(birds_dispersal_distance, birds_behavior), fill = factor(birds_dispersal_distance)), alpha = 0.3,colour=NA, show.legend = FALSE) +
  scale_color_viridis_d() +
  facet_wrap(~ birds_behavior, ncol = 4) + 
  labs(color = "Dispersal\n distance") +
  theme_bw() + xlab("Habitat Patch Size") + ylab("Shannon Diversity") +
  theme(
    legend.position = c(0.9,0.8)
    )

ggsave("Figures/AcrossBehav_Shannon_Dispersal_Lambda2_regular.png",width=8,height=6,units="in",dpi=600)

## Not fragmented communities Lambda = 1.7 - 4 - No Selection communities

```{r NoSelection_SLOSS_lambda14_notfrag,echo=FALSE,message=FALSE,warning=FALSE,eval=FALSE}

# pf = probability of fragmentation, hps = habitat patch size , dd=dispersal distance 
# hps = habitat patch size
# dd = dispersal distance 
# brb = bird rate birds 
nl@experiment <- experiment(expname="NoSelection_pf00_nofrag_dd1-3_brb1-4_regular",
                            outpath=outpath,
                            repetition=1,
                            tickmetrics="false",
                            idsetup="setup",
                            idgo="go",
                            runtime=0,
                            metrics=c("birds-at-deforestation", "calc-number-of-species","calc-shannon-diversity","count birds"),
                            variables = list("birds-dispersal-distance" = list(values=c(1.1, 1.2, 1.5, 2, 3)),
                                              "birth-rate-birds" = list(values=c(1.7, 2, 3, 4))
                                             ),
                            constants = list("world-width" = 199,
                                             "world-height" = 199,
                                             "initial-population" = 0,
                                             "max-birds-species" = 100,
                                             "death-rate-birds" = 1,
                                             "migration-rate-birds" = 0.0001,
                                             "Video" = "false",
                                             "prob-frag" =  0.6,
                                             "Deforestation-at-time" = 0,
                                             "end-time" = 1500,
                                             "Migration0-at-deforestation" = "false",
                                             "deforestation-type" = "\"regular\"",
                                             "birds-behavior" = "\"NoSelection\"",
                                             "replacement-rate" = 0.3
                                             ))

## Read Simulations NoFragmentation No Selection  regular Patterns

```{r readSimulationsNoFragNSelectionRegular, echo=FALSE,message=FALSE,warning=FALSE,eval=FALSE}
# Read previous simulations
#
if( file.exists("Simulations/sloss_sim_ff.rds")) {

# Read netlogo simulations and 
# Convert the list of mean_free_path to degraded_mfp and non_degraded_mfp
#
mdl <- read_sloss_simul(outpath,"NoSelection_pf00_nofrag_dd1-3_brb1-4_regular_ff.csv") %>% 
  mutate(habitat_patch_size = world_width, max_patch = (world_width+1) * (world_height+1) )

## Not fragmented communities Lambda = 1.7 - 4 - Hierarchical communities

```{r Hierarchical_SLOSS_lambda14_notfrag,echo=FALSE,message=FALSE,warning=FALSE,eval=FALSE}

# pf = probability of fragmentation, hps = habitat patch size , dd=dispersal distance 
# hps = habitat patch size
# dd = dispersal distance 
# brb = bird rate birds 
nl@experiment <- experiment(expname="Hierarchical_pf00_nofrag_dd1-3_brb1-4_regular",
                            outpath=outpath,
                            repetition=1,
                            tickmetrics="false",
                            idsetup="setup",
                            idgo="go",
                            runtime=0,
                            metrics=c("birds-at-deforestation", "calc-number-of-species","calc-shannon-diversity","count birds"),
                            variables = list("birds-dispersal-distance" = list(values=c(1.1, 1.2, 1.5, 2, 3)),
                                              "birth-rate-birds" = list(values=c(1.7, 2, 3, 4))
                                             ),
                            constants = list("world-width" = 199,
                                             "world-height" = 199,
                                             "initial-population" = 0,
                                             "max-birds-species" = 100,
                                             "death-rate-birds" = 1,
                                             "migration-rate-birds" = 0.0001,
                                             "Video" = "false",
                                             "prob-frag" =  0.6,
                                             "Deforestation-at-time" = 0,
                                             "end-time" = 1500,
                                             "Migration0-at-deforestation" = "false",
                                             "deforestation-type" = "\"regular\"",
                                             "birds-behavior" = "\"Hierarchical\"",
                                             "replacement-rate" = 0.3
                                             ))

## Read Simulations Not Fragmented  Hierarchical = habitat_patch_size == 200

```{r readSimulationsNoFragHierarchical, echo=FALSE,message=FALSE,warning=FALSE,eval=FALSE}
# Read previous simulations
#
if( file.exists("Simulations/sloss_sim_ff.rds")) {

# Read netlogo simulations and 
# Convert the list of mean_free_path to degraded_mfp and non_degraded_mfp
#
mdl <- read_sloss_simul(outpath,"Hierarchical_pf00_nofrag_dd1-3_brb1-4_regular_ff.csv") %>% 
  mutate(habitat_patch_size = world_width, max_patch = (world_width+1) * (world_height+1) )

#mdl <- sloss_sim %>% filter( Deforestation_at_time==0)

## Not fragmented communities Lambda = 1.7 - 4 - Selection communities

```{r BirthSelection_SLOSS_lambda14_notfrag,echo=FALSE,message=FALSE,warning=FALSE,eval=TRUE}

# pf = probability of fragmentation, hps = habitat patch size , dd=dispersal distance 
# hps = habitat patch size
# dd = dispersal distance 
# brb = bird rate birds 
nl@experiment <- experiment(expname="BirthSelection_pf00_nofrag_dd1-3_brb1-4_regular",
                            outpath=outpath,
                            repetition=1,
                            tickmetrics="false",
                            idsetup="setup",
                            idgo="go",
                            runtime=0,
                            metrics=c("birds-at-deforestation", "calc-number-of-species","calc-shannon-diversity","count birds"),
                            variables = list("birds-dispersal-distance" = list(values=c(1.1, 1.2, 1.5, 2, 3)),
                                              "birth-rate-birds" = list(values=c(1.7, 2, 3, 4))
                                             ),
                            constants = list("world-width" = 199,
                                             "world-height" = 199,
                                             "initial-population" = 0,
                                             "max-birds-species" = 100,
                                             "death-rate-birds" = 1,
                                             "migration-rate-birds" = 0.0001,
                                             "Video" = "false",
                                             "prob-frag" =  0.6,
                                             "Deforestation-at-time" = 0,
                                             "end-time" = 1500,
                                             "Migration0-at-deforestation" = "false",
                                             "deforestation-type" = "\"regular\"",
                                             "birds-behavior" = "\"BirthSelection\"",
                                             "replacement-rate" = 0.3
                                             ))

## Read Simulations Not Fragmented  Hierarchical = habitat_patch_size == 200

```{r readSimulationsNoFragBirthSelection, echo=FALSE,message=FALSE,warning=FALSE,eval=FALSE}
# Read previous simulations
#
if( file.exists("Simulations/sloss_sim_ff.rds")) {

# Read netlogo simulations and 
# Convert the list of mean_free_path to degraded_mfp and non_degraded_mfp
#
mdl <- read_sloss_simul(outpath,"BirthSelection_pf00_nofrag_dd1-3_brb1-4_regular_ff.csv") %>% 
  mutate(habitat_patch_size = world_width, max_patch = (world_width+1) * (world_height+1) )

#mdl <- sloss_sim %>% filter( Deforestation_at_time==0)

### 2023-08-22 ms_SLOSS.Rmd
Update manuscript

As global biodiversity is rapidly declining due to habitat loss, it is important to determine how to protect it. Still, the policies of many conservation agencies are prioritizing the preservation of a single large habitat area (SL) vs several small areas of the same size (SS), besides empirical support favoring SS. However, to date, most of the theories regarding this issue were based on single-species models. The aim of this work is to explore spatial multi-species models to verify the SS > SL pattern and determine the causes of it. We used three types of models: a classical neutral, neutral with habitat selection, and a competitive hierarchical model.  All models have three parameters, extinction, reproduction, and dispersal distance. In the classical neutral model, species are equivalent and have the same parameters, when there is a reproduction event they send the propagule to a surrounding patch blindly, if the patch is already occupied the propagule dies. In the neutral-with-habitat selection model, the propagules select an empty patch so their survival is ensured. In the hierarchical model, species have competitive dominance and more dominant ones can replace the less dominant with a replacement rate parameter. We use 100 species and start simulating the colonization of an empty area then we destroy a fixed amount of habitat using different regular patch sizes. The results for all models are almost always that SS retains more species than SL. The extinction of species in patches depends on the quotient between reproduction and extinction rates in relation to patch size. The only case when SL > SS is when dispersal is not enough to connect patches, and patch size is smaller than the minimum to prevent extinction. If patch size is above the critical size to maintain local populations the SS > SL pattern is due to the sampling effect of aggregated species distribution combined with the reduction of dispersal ability of species. High dispersal rates produce homogenization of species distribution and a reduction of species richness and this is why habitat fragmentation by habitat loss retains more species when patches are smaller. A prediction of the model is that if a set of species have more dispersal capabilities the SS effect would be lower than for species with less dispersal.

@@

Global biodiversity is rapidly declining due to habitat loss (quotes). Moreover, habitat fragmentation occurs at the same time with habitat loss, and the effects of both processes are difficult to disentangle. Landscape connectivity and organism mobility as a critical feature for population persistence [@Taylor1993]. Starting from a landmark paper [@Diamond1975], conservation biologists engaged in the challenging task of recommending protection measures given the environmental crisis coupled with lack of knowledge of most of the world’s biodiversity. Early recommendations claimed to follow ecological theory, in particular island biogeography theory to derive simple thumb rules decisions in order to better design protected areas. However, as noted rapidly, species/area relationships do not have a clear association with fragmentation per se but only with overall habitat loss [@Simberloff1982].
Aiming to design better management practices, this has led to a nearly 50-year Single Large vs Several Small (SLOSS) debate. While some authors argue that habitat loss is the most overwhelming effect on species loss, for others the complex factors arising from fragmentation further enhance species extinctions [@Fahrig2017;@Fletcher2018].

The SLOSS debate has been an important academic and practical discussion among conservation biology. In practice, a series of decision-maker targeted recommendations were proposed and remain rather uncritically in several conservation instances (quotes).
While the overall habitat loss is widely accepted as a process leading to long-term biodiversity loss (quotes), the propose that single large reserves retain more species that several small reserves of the same area (SL > SS) remain at a similar hierarchical state, even after several criticisms (reviewed in  @Fahrig2017). Moreover, empirical observational and experimental data shows the predominance of the opposite pattern (SS > SL) ever since the 80s and this fact has been reinforced to this day [@Simberloff1982; @Fahrig2020; @Hammill2020], add more. The actual processes driving to it remains elusive and it is probably multi-causal (quotes). In that scenario, model simulations could produce useful predictions targeting the relevant variables in order to better understand the processes leading to non-intuitive patterns.

The SLOSS debate has been approached from several theoretical frameworks, but most of them were based on single-species models. Moreover, spatial dimension in several works has been neglected as mere noise around long-term equilibrium dynamics (quotes, but see @Rybicki2013). In that context, spatial-explicit multi-species modeling could be meaningful as several organisms retain dispersal distance and other dispersal-related life traits far consistently among species (quotes).

@@

The environment comprises a 2D grid of patches that can be either suitable habitat or degraded. Bird agents have state variables for species identity, size, and color. Globals contain key parameters.

@@

The model implements stochasticity in multiple processes including dispersal distance, loss patterns, migration, and species selection. Emergent macro-level diversity and abundance patterns result from individual-level bird behaviors.

@@

* **MultipleContactNeutralMigrat.nlogo** A multiple species neutral/hierarchical model described in with 
	degraded sites that cannot be colonized

@@

Regular habitat patterns: squares of a range of sizes are distributed regularly over the simulation grid. The side of the squares where 4,20,30,63 sites.

Random block habitat patterns: habitat blocks of the same size are randomly distributed over the simulation grid. The sizes are identical to the used for regular habitat simulations. This habitat simulation results in patches that have a minimum size equal to the block size but a maximum size that is bigger, so I characterize this with the maximum patch size.

@@

![Relationship between number of species and habitat patch size for varying dispersal distances. Simulations were conducted using a model with 60% habitat destruction and regular habitat patterns, where all habitat patches are of equal size.](Figures/Species_Habitat_Dispersal_Lambda_regular.png)

![Relationship between Shannon Diversity and habitat patch size for varying dispersal distances. Simulations were conducted using a model with 60% habitat destruction and regular habitat patterns, where all habitat patches are of equal size.](Figures/Shannon_Habitat_Dispersal_Lambda_regular.png)

![Relationship between number of species and maximum patch size for varying dispersal distances. Simulations were conducted using a model with 60% habitat destruction and random block habitat patterns, where habitat blocks of the same size are randomly distributed over the simulation grid (i.e., the maximum patch size).](Figures/Species_MaxPatch_Dispersal_Lambda_block.png)

![Relationship between Shannon Diversity and maximum patch size for varying dispersal distances. Simulations were conducted using a model with 60% habitat destruction and random block habitat patterns, where habitat blocks of the same size are randomly distributed over the simulation grid (i.e., the maximum patch size).](Figures/Shannon_MaxPatch_Dispersal_Lambda_block.png)

@@

Overall, our results confirm the previous found pattern and nearly all models retained more species under the SS rather than the SL scenario. In both cases, the loss of species is relevant when habitat loss takes place, and no significant extinction biased towards SS occurs as a long-term dynamic. The extinction of species in patches depends on the quotient between reproduction and extinction rates in relation to patch size, i.e., the ability of a certain patch to be a source of propagules. The only case when SL > SS is when dispersal is not enough to connect patches, and size is smaller than the minimum to prevent patch extinction, as predicted elsewhere [@Rybicki2013]. Letting aside those cases (i.e., not reaching minimum dispersal distance and intra-patch sink dynamics), our model was able to produce the patterns observed so far in field studies [@Fahrig2020]. In that sense, we can hypothesize that our model variables (i.e., dispersal ability, patch size and lambda) may be biologically meaningful and enough to explain why SS retains more species than SL. Importantly, the spatial arrangement of available patches was not necessary to produce the spatial autocorrelation of occupied patches known to be responsible at least in part of the patterns observed in nature [@Fahrig2021]. In contrast, high dispersal rates produce homogenization of species distribution and a reduction of species richness, especially if high dispersal species are also good competitors. In that scenario, SS and SL are approximately equivalent for species diversity retention. A prediction of the model is that if a set of species have more dispersal capabilities the SS effect would be smaller than for species with less dispersal abilities.

Our model further illustrate that available patches are not always re-occupied following fragmentation even if the minimum size is met, therefore reinforcing the fact that dispersal is the main driver of the pattern, and that chance is involved in the final species pool when fragmentation takes place. Empirical data in large carnivores, for instance, support that several available patches are not occupied after isolation [@Paviolo2016].

@@

We acknowledge the limitations of our study, and consider it a strength that we have identified them in this initial model. This allows us to continue working on them in future research.

>>>>

As global biodiversity is rapidly declining due to habitat loss, it is important to determine how to protect it. Still, the policies of many conservation agencies are prioritizing the preservation of a single large habitat area (SL) vs several small areas of the same size (SS), besides empirical support favouring SS. However, to date, most of the theories regarding this issue were based on single-species models. The aim of this work is to explore spatial multi-species models to verify the SS > SL pattern and determine the causes of it. We used three types of models: a classical neutral, neutral with habitat selection, and a competitive hierarchical model.  All models have three parameters, extinction, reproduction, and dispersal distance. In the classical neutral model, species are equivalent and have the same parameters, when there is a reproduction event they send the propagule to a surrounding patch blindly, if the patch is already occupied the propagule dies. In the neutral-with-habitat selection model, the propagules select an empty patch so their survival is ensured. In the hierarchical model, species have competitive dominance and more dominant ones can replace the less dominant with a replacement rate parameter. We use 100 species and start simulating the colonization of an empty area then we destroy a fixed amount of habitat using different regular patch sizes. The results for all models are almost always that SS retains more species than SL. The extinction of species in patches depends on the quotient between reproduction and extinction rates in relation to patch size. The only case when SL > SS is when dispersal is not enough to connect patches, and patch size is smaller than the minimum to prevent extinction. If patch size is above the critical size to maintain local populations the SS > SL pattern is due to the sampling effect of aggregated species distribution combined with the reduction of dispersal ability of species. High dispersal rates produce homogenization of species distribution and a reduction of species richness and this is why habitat fragmentation by habitat loss retains more species when patches are smaller. A prediction of the model is that if a set of species have more dispersal capabilities the SS effect would be lower than for species with less dispersal.

@@

Global biodiversity is rapidly declining due to habitat loss (quotes). Moreover, habitat fragmentation occurs at the same time as habitat loss, and the effects of both processes are difficult to disentangle. Landscape connectivity and organism mobility as a critical feature for population persistence [@Taylor1993]. Starting from a landmark paper [@Diamond1975], conservation biologists engaged in the challenging task of recommending protection measures given the environmental crisis coupled with a lack of knowledge of most of the world’s biodiversity. Early recommendations claimed to follow ecological theory, in particular island biogeography theory to derive simple thumb rules decisions in order to better design protected areas. However, as noted rapidly, species/area relationships do not have a clear association with fragmentation per se but only with overall habitat loss [@Simberloff1982].
Aiming to design better management practices, this has led to a nearly 50-year Single Large vs Several Small (SLOSS) debate. While some authors argue that habitat loss is the most overwhelming effect on species loss, for others the complex factors arising from fragmentation further enhance species extinctions [@Fahrig2017;@Fletcher2018].

The SLOSS debate has been an important academic and practical discussion in conservation biology. In practice, a series of decision-maker targeted recommendations were proposed and remain rather uncritically in several conservation instances (quotes).
While the overall habitat loss is widely accepted as a process leading to long-term biodiversity loss (quotes), the proposition that single large reserves retain more species than several small reserves of the same area (SL > SS) remain at a similar hierarchical state, even after several criticisms (reviewed in  @Fahrig2017). Moreover, empirical observational and experimental data shows the predominance of the opposite pattern (SS > SL) ever since the 80s and this fact has been reinforced to this day [@Simberloff1982; @Fahrig2020; @Hammill2020], add more. The actual processes driving to it remain elusive and it is probably multi-causal (quotes). In that scenario, model simulations could produce useful predictions targeting the relevant variables in order to better understand the processes leading to non-intuitive patterns.

The SLOSS debate has been approached from several theoretical frameworks, but most of them were based on single-species models. Moreover, spatial dimension in several works has been neglected as mere noise around long-term equilibrium dynamics (quotes, but see @Rybicki2013). In that context, spatial-explicit multi-species modelling could be meaningful as several organisms retain dispersal distance and other dispersal-related life traits far consistently among species (quotes).

@@

The environment comprises a 2D grid of patches that can be either suitable habitat or degraded. Bird agents have state variables for species identity, size, and colour. Globals contain key parameters.

@@

The model implements stochasticity in multiple processes including dispersal distance, loss patterns, migration, and species selection. Emergent macro-level diversity and abundance patterns result from individual-level bird behaviours.

@@

* **MultipleContactNeutralMigrat.nlogo** A multiple species neutral/hierarchical model described in with 
        degraded sites that cannot be colonised

@@

Regular habitat patterns: squares of a range of sizes are distributed regularly over the simulation grid. The side of the squares were 4,20,30,63 sites.

Random block habitat patterns: habitat blocks of the same size are randomly distributed over the simulation grid. The sizes are identical to those used for regular habitat simulations. This habitat simulation results in patches that have a minimum size equal to the block size but a maximum size that is bigger, so I characterize this with the maximum patch size.

@@

![Relationship between number of species and habitat patch size for different neutral models and varying dispersal distances. Simulations were performed using a neutral model with a constant migration rate (0.0001). Following 200 time steps of the model, 60% of the habitat was destroyed, rendering it unavailable for the species. The destruction of habitat resulted in the formation of regular patterns, where all remaining patches exhibited equal sizes. Within the neutral models, three distinct approaches were employed. In the "No selection" model, juveniles dispersed to a random site within a distance determined by the dispersal kernel. If the site was occupied or non-habitat, they perished. In the hierarchical model, the key difference was that if a site was occupied, a superior competitor could replace an inferior competitor with a probability of replacement (0.3). Lastly, the "birth selection" model is identical the "No selection" model, except that juveniles selected an unoccupied habitable site within a maximum distance dictated by the dispersal kernel.](Figures/AcrossBehav_Species_Dispersal_Lambda2_regular.png)

![Relationship between Shannon Diversity and habitat patch size for different neutral models and varying dispersal distances. Simulations were performed using a neutral model with a constant migration rate (0.0001) and relationship of growth rate over death rate of 2. Following 200 time steps of the model, 60% of the habitat was destroyed, rendering it unavailable for the species. The destruction of habitat resulted in the formation of regular patterns, where all remaining patches exhibited equal sizes. Within the neutral models, three distinct approaches were employed. In the "No selection" model, juveniles dispersed to a random site within a distance determined by the dispersal kernel. If the site was occupied or non-habitat, they perished. In the hierarchical model, the key difference was that if a site was occupied, a superior competitor could replace an inferior competitor with a probability of replacement (0.3). Lastly, the "birth selection" model is identical the "No selection" model, except that juveniles selected an unoccupied habitable site within a maximum distance dictated by the dispersal kernel.](Figures/AcrossBehav_Shannon_Dispersal_Lambda2_regular.png)

@@

Overall, our results confirm the previously found pattern and nearly all models retained more species under the SS rather than the SL scenario. In both cases, the loss of species is relevant when habitat loss takes place, and no significant extinction biased towards SS occurs as a long-term dynamic. The extinction of species in patches depends on the quotient between reproduction and extinction rates in relation to patch size, i.e., the ability of a certain patch to be a source of propagules. The only case when SL > SS is when dispersal is not enough to connect patches, and size is smaller than the minimum to prevent patch extinction, as predicted elsewhere [@Rybicki2013]. Letting aside those cases (i.e., not reaching minimum dispersal distance and intra-patch sink dynamics), our model was able to produce the patterns observed so far in field studies [@Fahrig2020]. In that sense, we can hypothesize that our model variables (i.e., dispersal ability, patch size and lambda) may be biologically meaningful and enough to explain why SS retains more species than SL. Importantly, the spatial arrangement of available patches was not necessary to produce the spatial autocorrelation of occupied patches known to be responsible at least in part of the patterns observed in nature [@Fahrig2021]. In contrast, high dispersal rates produce homogenization of species distribution and a reduction of species richness, especially if high dispersal species are also good competitors. In that scenario, SS and SL are approximately equivalent for species diversity retention. A prediction of the model is that if a set of species have more dispersal capabilities the SS effect would be smaller than for species with less dispersal abilities.

Our model further illustrates that available patches are not always re-occupied following fragmentation even if the minimum size is met, therefore reinforcing the fact that dispersal is the main driver of the pattern, and that chance is involved in the final species pool when fragmentation takes place. Empirical data in large carnivores, for instance, support that several available patches are not occupied after isolation [@Paviolo2016].

@@

We acknowledge the limitations of our study and consider it a strength that we have identified them in this initial model. This allows us to continue working on them in future research.
